---
title: "Dataset manipulation pt 2"
author: "Amanda Mae Woodward"
date: "2024-10-15"
output: html_document
---

*It is okay to turn in assignment late, make sure to turn it in before Tuesday class *

We're going to use this part of class to review how to do the learning outcomes from the dataset manipulation notes. 

### load data
1. Open the palmerpenguins library
```{r}

# Type in the console 'library(palmerpenguins)'

library(palmerpenguins)

```

2. Open the penguins dataset
```{r}
data("penguins")
```

3. Explore the raw dataset. What can you find out about it? 

The data looks at the penguin body size, clutch, and blood sample.
- Penguin data is a subset of penguins_raw
- Penguin data looks better because its more clear  

4. Are the data in long or wide format? 
```{r}
# The penguin data is a long format because multiple penguins have more than one row.  

# How to figure it out? We can look at the species, sample number, indidviaul ID

# Another way to look at the summary of the column 'sample number'
summary(penguins_raw$`Sample Number`)
summary(as.factor(penguins_raw$`Sample Number`))

# For species 
summary(as.factor(penguins_raw$Species))

# For ID 
summary(as.factor(penguins_raw$`Individual ID`)) # This is what we'll be using in class because it looks at the individual penguins 
```

5. let's practice going from one format to the other. 

6. create two datsets. One should have the individual id, region, and island. The other should have individual id, species, flipper length, culmen length, culmen depth, and body mass. 

```{r}
# We will be using flipper length 

library(tidyverse)

tinyPenguins <- penguins_raw %>% select("Individual ID", "Sample Number", "Flipper Length (mm)")

# Notes: individual id is not nested within sample number. We can still run this code, but its going to be messy. 


# Change data tinyPenguin from long to wide format 
penguinWide <- pivot_wider(tinyPenguins, id_cols= "Individual ID", names_from= "Sample Number", values_from= "Flipper Length (mm)")

```

```{r}

# make a long data longer 
penguinLongData <- penguins_raw %>% pivot_longer(cols= c("Culmen Length (mm)", "Culmen Depth (mm)", "Flipper Length (mm)", "Body Mass (g)"), names_to= "bodyMeasure", values_to= "measurements")

# make the longer data to wide 
penguinWide2  <- penguinLongData %>% pivot_wider(id_cols= c("Individual ID", "Sample Number"), names_from= "bodyMeasure", values_from= "measurements")

# When we run, we get the Individual ID and Sample Number column in the front of the data. The body measure are the columns and the measurement are the values. 
```

7. Join the two datasets you made together.
```{r}
# Data of the first part 
penguinA <- penguins_raw %>% select("Individual ID", "Region","Island")


# Data of the second part
penguinB <- penguins_raw %>% select("Individual ID", "Culmen Length (mm)", "Culmen Depth (mm)", "Flipper Length (mm)", "Body Mass (g)")


```

8. Separate the species column into the common name and the scientific name 
```{r}
a <- left_join(penguinA, penguinB)

# This combine the two data set 

# To seperate penguin_raw
b <- separate(penguins_raw, col = "Species", into= c("commonName", "scientificName"), sep = "")
```
9. change the levels of clutch completion to indicate whether the penguin has two eggs or does not have two eggs
```{r}
penguins_raw$`Clutch Completion` 

penguins_raw$`Clutch Completion` <- as.factor(penguins_raw$`Clutch Completion`)
levels(penguins_raw$`Clutch Completion`) <- c("not two eggs", "two eggs")
penguins_raw$`Clutch Completion`

```

10. create a column that has body mass in kilograms.

```{r}
penguins_raw$`bodyMass(kg)` <- NA
```

```{r}
penguins_raw$`bodyMass(kg)` <- penguins_raw$`Body Mass (g)` /1000

```

11. copy island to a new column. Recode this variable so that each island is renamed to be just the first initial

```{r}
penguins_raw$`island2` <- NA
```

```{r}
penguins_raw$`island2` <- penguins_raw$Island 

library(car)
summary(as.factor(penguins_raw$island2))
penguins_raw$island2 <- car:: recode(penguins_raw$island2, recodes= "'Biscoe'='B'; 'Dream'='D';'Torgersen'='T'", separator= ";")
```

