---
title: "ggplot part 2"
author: "Amanda Mae Woodward"
date: "2024-10-22"
output: html_document
---

# Learning Outcomes
By the end of this class, students should be able to: 
- Explain how to alter information in the legend 
- Describe how to use facet wrap
- Explain how to export graphs


### Learning Outcome 1: Explain how to alter information in the legend
It can also be helpful to understand how to modify your legend: 
`+ theme(legend.position="none")` will remove the legend. You can also choose to move the legend to the top or bottom of the graph

Let's try this on a graph from the diamonds dataset: 
Create a subset of the diamonds dataset:
```{r}
library(tidyverse)
data(diamonds)

# How to remove the legend 
ggplot(diamonds, aes(carat, price, color=cut))+
  geom_point() +
  theme(legend.position= "none")
```

We can also change the location: 
```{r}
# How to move the legend to the bottom of the graph 
ggplot(diamonds, aes(carat, price, color=cut))+
  geom_point() +
  theme(legend.position= "bottom")
```

To Change the legend title, you can use a couple of different codes, depending on the type of data: 
`scale_color_discrete(name="Title")` works for scatter plots filled in with a factor
`scale_fill_discrete(name="Title)` should work for a bar graph filled in by a factor
`scale_color_continuous()` and `scale_fill_continuous()` should work for a numeric fill
```{r}
# How to change the names in the legend 
ggplot(diamonds, aes(carat, price, color=cut))+
  geom_point() +
  theme(legend.position= "bottom")
```

### Practice:

1) make a graph of disp and hp from mtcars. Include miles per gallon as a third variable
```{r}
data(mtcars)
ggplot(mtcars, aes(disp, hp, color=mpg))+ 
  geom_point()


# Use discrete = FALSE because the variable is continuous 
```

2) Move the legend to the top of the graph
```{r}
data(mtcars)
ggplot(mtcars, aes(disp, hp, color=mpg))+ 
  geom_point()+ 
  theme_classic()+
  theme(legend.position= "top")
```

3) Remove the legend from the graph 
```{r}
data(mtcars)
ggplot(mtcars, aes(disp, hp, color=mpg))+ 
  geom_point()+
  theme_classic()+
  theme(legend.position= "none")
```

4) Change the title of the legend
```{r}
data(mtcars)
ggplot(mtcars, aes(disp, hp, color=mpg))+ 
  geom_point()+ 
  theme_classic()+
  scale_color_viridis(discrete= FALSE, option="H", name= "Displacement and Horse Power")
```

### Learning Outcome 2: Describe how to use facet wrap
Facet Wrap allows us to incorporate a third variable in another way. Specifically, it panels graphs so that we can make an individual graph for each level of a factor.

we need to make a graph and save it (we'll use carat and price)
```{r}
mpgByCyl <- mtcars %>%
  group_by(cyl)%>%
  summarize(mean = mean(mpg))

ggplot(mpgByCyl, aes(cyl, mean)) + 
  geom_bar(stat="identity")

mpg <- ggplot(mtcars, aes(mpg))+
  geom_bar() 
mpg+facet_wrap(~cyl)
```

```{r}

# Histogram 
ggplot(mtcars, aes(mpg))+ 
  geom_histogram(bins= 5, fill="blue")+
  facet_wrap(~cyl)+
  ylab(label= "frequency")+
  xlab(label="miles per gallon")
  
```
then we'll use cut to create the facets wrap graph
```{r}
# option 1
labels<- c("4"= "4 cylinder", "6"= "6 cylinder", "8"= "8 cylinder")

ggplot(mtcars, aes(mpg))+ geom_histogram(bins= 5, fills= "blue")+  facet_wrap(~as.factor(cyl), labeller= as_labeller(labels))

# option 2
mtcars$cyl <- as.factor(mtcars$mpg)
levels(mtcars$cyl) <- c("4 cylinders", "6 cylinders", "8 cylinders")
ggplot(mtcars, aes(mpg))+ geom_histogram()+facet_wrap(~cyl)
```

```{r}

```

### Practice: 
1) create a facet wrap graph to generate a scatterplot of mpg and hp based on the number of cylinders a car has
```{r}
ggplot(mtcars, aes(mpg, hp))+
  geom_point()+
  facet_wrap(~cyl, labeller= as_labeller(labels))+
  theme_classic()+
  ylab(label= "horse power")+
  xlab(label= "miles per gallon")
```

```{r}
# changing the font and size 
prettyGraph <- ggplot(mtcars, aes(mpg, hp))+
  geom_point()+
  facet_wrap(~cyl, labeller= as_labeller(labels))+
  theme_classic()+
  ylab(label= "horse power")+
  xlab(label= "miles per gallon")+
  scale_color_viridis(discrete= FALSE, option = "D")

finalPretty <- prettyGraph + theme(legend.position= "none", text = element_text(family = "Times New Roman", size= 18))

```

### Learning Outcome 3: Explain how to export graphs
to export graphs, I use the command `ggsave()`
Specifically, you can use the following arguments: 
`ggsave("name you want to give the graph", width, height, units, dpi)`
```{r}

# How to save and export your graph on the computer 

ggsave("PSY4802-FinalPrettyGraph.tiff", width= 10, height=10, units="in", dpi= 300)

# dpi is the resolution 

# Save without project by using *getwd()*
```
I don't usually specify a path in ggsave, but make sure my r code is in the directory I'd like it to be saved in. you can check your working directory by using this code
```{r}


```
you can change your working directory using this code: 
```{r}


```

