---
title: "Best and Worst Graph"
author: "Kasia"
date: "2024-10-23"
output: html_document
---

```{r}
# import data 
candyData <- read.csv("C:/Users/vang3434/Downloads/candy-data-1.csv")

```
For binary variables, 1 means yes, 0 means no. The data contains the following fields:

- chocolate: Does it contain chocolate?
- fruity: Is it fruit flavored?
- caramel: Is there caramel in the candy?
- peanutalmondy: Does it contain peanuts, peanut butter or almonds?
- nougat: Does it contain nougat?
- crispedricewafer: Does it contain crisped rice, wafers, or a cookie component?
- hard: Is it a hard candy?
- bar: Is it a candy bar?
- pluribus: Is it one of many candies in a bag or box?
- - sugarpercent: The percentile of sugar it falls under within the data set.
- pricepercent: The unit price percentile compared to the rest of the set.
- winpercent: The overall win percentage according to 269,000 matchups.

Your graph must include information about *sugar percent* and *price percent*. All other decisions are up to you as the person making the graph! 

# Best Graph

```{r}
library(tidyverse)
```

```{r}
library("Rmisc")
```

```{r}
ggplot(candyData, aes(sugarpercent, pricepercent, color=as.factor(chocolate)))+ 
  geom_point()+ 
  labs(x= "Percentile of Sugar", y= "Unit Price Percentile", title= "Halloween Candy Data")+ 
  theme_light()+
  geom_smooth(method= "lm")+
  scale_color_manual(values=c("black", "orange"))
```


# Worst Graph 
```{r}
ggplot(candyData, aes(sugarpercent, pricepercent, color=as.factor(chocolate)))+ 
  geom_point(color="lightblue", size=19) +
  labs(x= "Percentile of Sugar", y="Unit Price Percentile", title= "Halloween Candy Data")+
  theme_minimal()+
  geom_jitter(color="skyblue", shape= 8, size=10)+
  geom_hline(yintercept = 0.5, color="grey", linetype= "dashed")+
  geom_vline(xintercept = 0.6, color= "grey")
  
```